---
title: "Module 0 — From Raw Files to a Biobank-Grade Sample Sheet"
subtitle: "BBCC2026 Tutorial: Organizing Metadata Before Any Analysis Begins"
author: |
  **Francesco Cecere**
  *Senior Bioinformatic Researcher*
  AORN Santobono-Pausilipon
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

<style>
body { font-size: 15px; }

h1 {
  border-bottom: 3px solid #2980b9;
  padding-bottom: 8px;
  color: #1a252f;
  margin-top: 32px;
}
h2 {
  border-bottom: 1px solid #aed6f1;
  padding-bottom: 4px;
  color: #2471a3;
  margin-top: 22px;
}
h3 { color: #1f618d; }

blockquote {
  background-color: #eaf4fd;
  border-left: 5px solid #2980b9;
  padding: 14px 20px;
  border-radius: 0 6px 6px 0;
  font-style: italic;
  color: #2c3e50;
  margin: 18px 0;
}
blockquote p { margin: 0; }

.value-boxes {
  display: flex;
  flex-wrap: wrap;
  gap: 16px;
  margin: 24px 0 32px 0;
}
.vbox {
  background-color: #2980b9;
  background: linear-gradient(135deg, #2980b9 0%, #1a5276 100%);
  color: #ffffff !important;
  border-radius: 10px;
  padding: 20px 28px;
  min-width: 150px;
  text-align: center;
  box-shadow: 0 3px 10px rgba(0,0,0,0.15);
  flex: 1;
  transition: transform 0.2s;
}
.vbox:hover { transform: translateY(-3px); }
.vbox.green  { background-color: #27ae60; background: linear-gradient(135deg, #27ae60 0%, #1e8449 100%); }
.vbox.red    { background-color: #e74c3c; background: linear-gradient(135deg, #e74c3c 0%, #a93226 100%); }
.vbox.purple { background-color: #8e44ad; background: linear-gradient(135deg, #8e44ad 0%, #6c3483 100%); }
.vbox .vbox-number, .vbox-number {
  font-size: 2.6em; font-weight: 700; line-height: 1.1; color: #ffffff !important;
}
.vbox .vbox-label, .vbox-label {
  font-size: 0.82em; margin-top: 5px; opacity: 0.92; letter-spacing: 0.02em; color: #ffffff !important;
}

.alert-sig {
  background: #d5f5e3;
  border-left: 5px solid #27ae60;
  padding: 12px 18px;
  border-radius: 0 6px 6px 0;
  margin: 12px 0;
}
.alert-notsig {
  background: #fdf2f8;
  border-left: 5px solid #a569bd;
  padding: 12px 18px;
  border-radius: 0 6px 6px 0;
  margin: 12px 0;
}
.alert-info {
  background: #fef9e7;
  border-left: 5px solid #f39c12;
  padding: 12px 18px;
  border-radius: 0 6px 6px 0;
  margin: 12px 0;
}

.dataTables_wrapper { margin: 10px 0 24px 0; font-size: 0.93em; }
table.dataTable thead th {
  background-color: #2980b9 !important;
  color: white !important;
}
.dt-buttons .btn { font-size: 0.82em; padding: 3px 10px; }

.report-footer {
  margin-top: 40px;
  padding-top: 16px;
  border-top: 1px solid #ccc;
  font-size: 0.85em;
  color: #777;
  text-align: center;
}

.tocify {
  background-color: #1a252f !important;
  border: none !important;
  border-radius: 8px;
  padding: 6px 0;
}
.tocify .tocify-item > a {
  color: #bdc3c7 !important;
  padding: 6px 14px;
  font-size: 0.88em;
  border-radius: 4px;
  transition: background 0.18s, color 0.18s;
  text-decoration: none;
}
.tocify .tocify-item > a:hover {
  background-color: #2c3e50 !important;
  color: #ffffff !important;
}
.tocify .tocify-item.active > a {
  background-color: #2980b9 !important;
  color: #ffffff !important;
  font-weight: 600;
  border-radius: 4px;
}
.tocify-subheader {
  background-color: #1a252f !important;
}
.tocify-subheader .tocify-item > a {
  padding-left: 26px !important;
  font-size: 0.82em !important;
  color: #7f8c8d !important;
}
.tocify-subheader .tocify-item.active > a {
  background-color: #21618c !important;
  color: #ecf0f1 !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  dpi       = 96
)
```

```{r libraries, include=FALSE}
library(tidyverse)   # dplyr, ggplot2, tidyr, readr, etc.
library(plotly)      # interactive plots
library(DT)          # interactive downloadable tables
library(knitr)       # kable for simple inline tables
library(scales)      # axis formatting
```

# Welcome to the BBCC2026 Tutorial

> This tutorial offers a hands-on introduction to how biological and clinical data are organized, analyzed, and turned into clinically meaningful findings. Starting from the organization of metadata and biobanking, we will follow a real dataset through an expression analysis to identify candidate biomarkers with downstream interpretation. Each module is connected back to rare disease and cancer biology from a pediatric point of view, highlighting what makes them challenging to analyze.

<div class="value-boxes">
<div class="vbox"><div class="vbox-number">4</div><div class="vbox-label">MODULES</div></div>
<div class="vbox green"><div class="vbox-number">45</div><div class="vbox-label">SAMPLES</div></div>
<div class="vbox purple"><div class="vbox-number">17,616</div><div class="vbox-label">GENES (RAW)</div></div>
<div class="vbox red"><div class="vbox-number">GSE303967</div><div class="vbox-label">SOURCE DATASET</div></div>
</div>

## Roadmap

| Module | Title | Focus |
|--------|-------|-------|
| 0 | Metadata & Biobanking | Turning raw file names into a traceable, biobank-grade sample sheet |
| 1 | Exploratory Analysis & QC | Library sizes, distributions, PCA, sample-sample correlation |
| 2 | Expression Analysis & Biomarker Discovery | Filtering, per-gene testing, volcano plot, candidate table |
| 3 | Downstream Interpretation | Curated pediatric-cancer gene sets, enrichment testing, clinical framing |

This module — **Module 0** — is deliberately the least "computational" and the most conceptual, because in real translational research **the analysis is only as trustworthy as the metadata behind it.** This is especially true in **pediatric rare disease and pediatric oncology**, where cohorts are small by definition, samples are precious and often irreplaceable, and a single mislabeled sample can invalidate months of work. Before touching a single p-value, we need to know: *which sample is which, which samples belong together, and what, exactly, are we comparing?*

**Session plan (≈50 minutes):** why metadata matters (10 min) → what good metadata looks like (10 min) → biobank basics (10 min) → hands-on: build & debug a mini biobank record (20-25 min) → plenary debrief (5 min) → closing pediatric lens (5 min) → applying all of this to the real GSE303967 dataset.

# Part 1 — Why Metadata Matters

> Data organization is not bureaucracy — it is what makes any later analysis trustworthy.

The fastest way to feel this, rather than just be told it, is to watch a perfectly ordinary-looking comparison quietly produce nonsense. Below is a tiny, six-patient toy cohort — the kind of table a well-meaning research assistant might type up in an afternoon — comparing an `age_recorded` value between two groups.

```{r broken-cohort-data, echo=TRUE}
# A small, intentionally ordinary-looking toy cohort. Nothing here looks
# obviously wrong at a glance.
broken_cohort <- tribble(
  ~patient_id, ~age_recorded, ~sex, ~group,
  "C-01",      3,             "F",  "Responder",
  "C-02",      5,             "M",  "Responder",
  "C-03",      4,             NA,   "Responder",
  "C-04",      36,            "F",  "Non-responder",
  "C-05",      24,            NA,   "Non-responder",
  "C-06",      6,             "M",  "Non-responder"
)
```

```{r table-broken-cohort}
kable(broken_cohort, caption = "A toy six-patient cohort. Nothing here is flagged as an error by R — the values are all valid numbers and strings.")
```

Now we run the single most ordinary thing anyone would do with this table: compare the two groups.

```{r broken-cohort-analysis, echo=TRUE}
# A completely standard group comparison — nothing unusual about this code.
age_comparison <- broken_cohort %>%
  group_by(group) %>%
  summarise(mean_age = mean(age_recorded), n = n(), .groups = "drop")
```

```{r table-broken-comparison}
kable(age_comparison, caption = "Mean age_recorded by group — computed correctly, from meaningless input.")
```

<div class="alert-info">
**The reveal.** `C-04` and `C-05` were recorded in **months** (36 and 24), while every other patient was recorded in **years**. There is no `units` column to warn us, so R has no way to know, and the code above ran without a single warning or error. The result — a "Non-responder" group that looks nearly six times older than the "Responder" group — is not a biological finding. It is a metadata failure wearing the costume of a result. **This is the entire point: R does not know what a number means, only what it is.**
</div>

A second, quieter failure is sitting in the same table: `sex` is missing for two of the six patients (`C-03`, `C-05`). Watch what happens to a routine sex-stratified summary:

```{r broken-cohort-sex, echo=TRUE}
# A routine stratified count — filter(!is.na(...)) is such a common, almost
# invisible habit that it is easy to not notice what it silently discards.
sex_summary <- broken_cohort %>%
  filter(!is.na(sex)) %>%
  count(sex)

cat("Patients in the full cohort:", nrow(broken_cohort), "\n")
cat("Patients included in the sex-stratified summary:", sum(sex_summary$n), "\n")
```

<div class="alert-info">
**The reveal.** One line of ordinary, idiomatic tidyverse code (`filter(!is.na(sex))`) silently dropped **2 of 6 patients — a third of the cohort** — with no warning printed anywhere. In a 500-patient adult study, losing 2 patients to a missing field is a rounding error. In a rare pediatric cohort of 6, 15, or 30 patients, it can be the difference between a result and no result at all — and if it happens silently, nobody even gets the chance to ask why.
</div>

**Garbage in, garbage out** is usually said as an abstraction. The two examples above are the concrete version: the code was correct in both cases. The *data* is what failed, and nothing in R's normal execution flags that failure for you. That is exactly why metadata organization is a first-class analytical skill, not paperwork that happens before "the real work" begins.

# Part 2 — What Good Metadata Looks Like

## Minimal Fields Every Biological Sample Needs

At a minimum, a usable biobank/sample record needs to answer: *who, what, when, and under what permission.*

| Field | Why it is non-negotiable |
|-------|---------------------------|
| **Patient/subject ID** | The single key linking a sample to its clinical context — must be unique, always. |
| **Sample type** | Tumor tissue, blood, saliva, etc. — determines what analyses are even valid. |
| **Collection date** | Anchors the sample in time relative to diagnosis, treatment, and the patient's age. |
| **Diagnosis** | Ideally from a controlled vocabulary, not free text (more on this below). |
| **Age at collection** | Needed for almost any biological interpretation — and, as Part 1 showed, needs an explicit, consistent unit. |
| **Consent status** | Whether the sample can legally and ethically be used for research at all — see Part 3. |

## Standards, Briefly

Two ideas are worth knowing exist, without a deep dive here:

- **FAIR principles** — data and metadata should be **F**indable, **A**ccessible, **I**nteroperable, and **R**eusable. In practice, for this tutorial, that mostly means: consistent IDs, documented units, and machine-readable fields rather than prose.
- **Ontologies**, such as the **Human Phenotype Ontology (HPO)** for clinical phenotypes, give every term a stable, shared identifier (e.g. an HPO code) instead of free text. "Kid kidney tumor" and "pediatric renal neoplasm" are the same thing to a clinician but look like two different diagnoses to a computer — an ontology term fixes that.

## Messy vs. Clean, Side by Side

Here is a miniature, self-contained illustration of the difference structure makes — not the exercise table you will work with shortly, just a quick contrast.

```{r messy-vs-clean, echo=TRUE}
# A minimal "messy" record: inconsistent casing, a merged free-text field,
# and a blank standing in for a missing value with no way to tell if that
# blank means "unknown" or "not applicable".
messy_example <- tribble(
  ~ID,  ~"Age/sex",  ~Notes,
  "s1", "4, girl",   "wilms",
  "s2", "unknown",   "",
  "s3", "6 yo M",    "neuroblastoma??"
)

# The same information, restructured: one field per variable, explicit NA
# for genuinely missing values, and a controlled diagnosis term.
clean_example <- tribble(
  ~patient_id, ~age_years, ~sex, ~diagnosis,
  "PT-9001",   4,          "F",  "Wilms tumor",
  "PT-9002",   NA,         NA,   NA,
  "PT-9003",   6,          "M",  "Neuroblastoma"
)
```

```{r table-messy}
kable(messy_example, caption = "Messy: mixed fields, inconsistent formatting, ambiguous blanks.")
```

```{r table-clean}
kable(clean_example, caption = "Clean: one column per variable, explicit NA, standardized terms.")
```

Nothing was lost going from messy to clean — the *information content* of `s2`/`PT-9002` and `unknown` is identical. What changed is that the clean version can be filtered, joined, counted, and audited by code; the messy version can only be read by a human, one row at a time.

# Part 3 — Biobank Basics

A biobank is, at its core, a system for keeping **three linked things** in sync for every sample, indefinitely:

1. **The physical sample** (tumor tissue, blood, DNA, RNA — stored, tracked, and eventually consumed by analyses like the one in this tutorial).
2. **The linked clinical/phenotypic data** (diagnosis, treatment, outcome — everything that gives the sample scientific meaning).
3. **The consent record** (what the donor agreed to, and for how long).

If any one of these three drifts out of sync with the other two — a sample outlives its consent, a clinical update never reaches the biobank record, a sample is used for a purpose broader than what was consented — the biobank has, functionally, failed at its one job, regardless of how well the freezer is organized.

<div class="alert-info">
**Pediatric consent is not a smaller version of adult consent — it is structurally different.** A child cannot give legally binding consent; a parent or guardian provides it on their behalf, often alongside the child's own **assent** (age-appropriate agreement, even without full legal capacity). Because childhood spans years of rapidly changing capacity to understand and decide, many pediatric biobanks require **re-consent** as the child grows — most concretely, from parental consent toward the now-adult individual's own consent once they reach the age of majority. A sample collected in infancy may need its consent status revisited a decade or more later, which is a governance problem that essentially does not exist in most adult biobanking.
</div>

# Part 4 — Hands-On: Build & Debug a Mini Biobank Record

**Groups of 3-4, 20-25 minutes.** Below is a working (if slightly expanded — 8 patients rather than 5-6, so every error type below gets a clean row of its own) biobank table for a fictional pediatric cohort. Two records are largely blank and need to be completed from short case notes; the rest are already filled in, but **each of the remaining rows hides exactly one deliberately seeded error** from a real, common biobank data-quality category. Note that patients are identified only by ID, never by name — that is itself a deliberate, realistic biobanking choice, not an oversight.

**Task:** as a group, (1) fill in the two incomplete records using the case notes, and (2) find and propose a fix for the error in each of the other six records. You do not need R for this part — read the table and notes, and write down your answers.

```{r mini-biobank-data, echo=TRUE}
# The working (messy) mini biobank table for the group exercise.
# Two rows (PT-2031, PT-2032's *first* appearance) are incomplete on purpose;
# the other rows each carry exactly one seeded data-quality error.
mini_biobank_messy <- tribble(
  ~patient_id, ~date_of_birth, ~age_years, ~sex, ~diagnosis,          ~collection_date, ~sample_type,   ~consent_status,
  "PT-2031",   "2024-01-03",   NA,         NA,   NA,                  NA,               NA,             NA,
  "PT-2032",   "2019-11-22",   NA,         NA,   NA,                  NA,               NA,             NA,
  "PT-2032",   "2016-06-30",   8,          "M",  "Ewing sarcoma",     "2024-09-10",     "Tumor tissue", "Consented",
  "PT-2033",   "2023-05-14",   18,         "F",  "Retinoblastoma",    "2024-11-02",     "Tumor tissue", "Consented",
  "PT-2034",   "2020-02-08",   5,          NA,   "Wilms tumor",       "2025-03-01",     "Tumor tissue", "Consented",
  "PT-2035",   "2022-07-19",   2,          "M",  "Neuroblastoma",     "2021-12-05",     "Tumor tissue", "Consented",
  "PT-2036",   "2015-09-27",   9,          "F",  "kid kidney tumor",  "2024-10-14",     "Tumor tissue", "Consented",
  "PT-2037",   "2018-12-01",   6,          "M",  "Osteosarcoma",      "2025-01-20",     "Tumor tissue", NA
) %>%
  mutate(date_of_birth = as.Date(date_of_birth), collection_date = as.Date(collection_date))
```

```{r table-mini-biobank}
datatable(
  mini_biobank_messy,
  extensions = "Buttons",
  options = list(dom = "Bfrtip", buttons = c("csv", "excel", "pdf"), pageLength = 10),
  rownames = FALSE,
  caption = "The mini biobank exercise table — 2 records to complete, 6 records each hiding one error."
)
```

## Case Notes (for the two incomplete records)

> **Referral note — record 1 (date of birth 3 January 2024).** Female infant. Presented with a palpable abdominal mass; imaging and biopsy were consistent with neuroblastoma. A tumor tissue sample was biobanked on 20 March 2025, after the parents signed the institutional consent form.

> **Referral note — record 2 (date of birth 22 November 2019).** Male child. Diagnosed with osteosarcoma of the left femur; a tumor biopsy sample was biobanked on 5 June 2024, following parental consent.

**Convention for this exercise:** record age at collection as **completed years** (i.e. round down), computed from date of birth and collection date — not estimated by eye.

## Reveal the Answer Key

Work out your group's answers first. Then expand the section below to compare.

<details>
<summary><strong>Click to reveal the answer key</strong></summary>

| Record | Category | What was wrong | Fix |
|--------|----------|-----------------|-----|
| PT-2031 | Incomplete | Age, sex, diagnosis, collection date, sample type, consent were blank | From the note: age 1 year (2024-01-03 → 2025-03-20), sex F, diagnosis Neuroblastoma, collection date 2025-03-20, sample type Tumor tissue, consent Consented |
| PT-2032 (1st row) | Incomplete | Same fields blank | From the note: age 4 years (2019-11-22 → 2024-06-05), sex M, diagnosis Osteosarcoma, collection date 2024-06-05, sample type Tumor tissue, consent Consented |
| PT-2032 (2nd row) | **Duplicate ID** | This is a *different* child (different DOB, sex, diagnosis) filed under the same ID as record 2 | Assign a new, unique ID (e.g. `PT-2038`) — never resolve a duplicate ID by deleting a row without checking which record, if either, is correct |
| PT-2033 | **Unit inconsistency** | `age_years = 18` — but the gap between DOB and collection date is 538 days (≈17.7 **months**, i.e. ≈1.5 years). Someone entered months into a years field. A diagnosis of retinoblastoma (typically <5 years old) should also have raised suspicion of an 18-year-old. | Correct value: `age_years = 1` (completed years) |
| PT-2034 | **Missing data** | `sex` is blank | Cannot be inferred from this table — flag for follow-up against the original clinical record. Do **not** guess. |
| PT-2035 | **Impossible value** | `collection_date` (2021-12-05) is *before* `date_of_birth` (2022-07-19) | Likely a transcription typo (2021 vs. 2024) — given `age_years = 2` and the DOB, the intended date was almost certainly in late 2024. Flag and verify against the source record rather than silently "correcting" the year. |
| PT-2036 | **Non-standardized terminology** | `diagnosis = "kid kidney tumor"` — free text, not a controlled term | Correct value: `"Wilms tumor"` |
| PT-2037 | **Missing data** | `consent_status` is blank | Cannot be inferred, and in real biobank governance this is a **stop condition**: the sample must not be used for research until consent is verified — this is not a value to fill in from guesswork. |

</details>

## From Manual Review to Automated Checks

Everything your group just did by eye can — and in a real pipeline, *should* — also be done programmatically, so it runs on every new batch of samples without relying on someone noticing by hand.

```{r automated-checks, echo=TRUE}
# 1. Duplicate IDs: any patient_id appearing more than once.
duplicate_ids <- mini_biobank_messy %>% count(patient_id) %>% filter(n > 1)

# 2. Impossible dates: collection before birth.
impossible_dates <- mini_biobank_messy %>%
  filter(!is.na(collection_date) & collection_date < date_of_birth)

# 3. Non-standardized diagnoses: anything not in a small controlled vocabulary.
standard_diagnoses <- c("Neuroblastoma", "Retinoblastoma", "Wilms tumor", "Ewing sarcoma", "Osteosarcoma")
nonstandard_diagnoses <- mini_biobank_messy %>%
  filter(!is.na(diagnosis) & !diagnosis %in% standard_diagnoses)

# 4. Implausible ages for a pediatric cohort: a simple sanity bound that also
#    happens to catch the years/months mix-up, since 18 "years" is implausible
#    here but 18 months would not have been.
implausible_ages <- mini_biobank_messy %>% filter(!is.na(age_years) & age_years > 17)

# 5 & 6. Missing sex / consent.
missing_sex     <- mini_biobank_messy %>% filter(is.na(sex))
missing_consent <- mini_biobank_messy %>% filter(is.na(consent_status))
```

```{r verify-automated-checks, echo=TRUE}
cat("Duplicate patient_id values:", paste(duplicate_ids$patient_id, collapse = ", "), "\n")
cat("Records with an impossible collection date:", paste(impossible_dates$patient_id, collapse = ", "), "\n")
cat("Records with a non-standardized diagnosis:", paste(nonstandard_diagnoses$patient_id, collapse = ", "), "\n")
cat("Records with an implausible age (>17 years):", paste(implausible_ages$patient_id, collapse = ", "), "\n")
cat("Records with missing sex:", paste(missing_sex$patient_id, collapse = ", "), "\n")
cat("Records with missing consent status:", paste(missing_consent$patient_id, collapse = ", "), "\n")
```

Every single seeded error from the answer key above is caught by five short `dplyr` filters — no custom function required. This is the real argument for structured metadata: it is not just easier to *read*, it is checkable **by a machine**, which a free-text spreadsheet never is.

# Part 5 — Plenary Debrief

A few questions worth putting to the room before moving on:

- **Which errors did your group spot fastest, and which took longest?** Duplicate IDs and impossible dates are usually caught almost immediately — they are internally contradictory and need no outside knowledge. Non-standardized terminology (`"kid kidney tumor"`) is typically the slowest, because catching it requires *already knowing* that a standard term exists — which is precisely the problem ontologies like HPO are built to solve.
- **Which errors could be directly fixed, and which could only be flagged?** The unit error and the terminology error had clear, computable fixes. The missing sex, missing consent, and impossible date did not — the honest fix was "go back to the source record," not "guess a plausible value." Knowing the difference is itself a data-quality skill.

# Part 6 — The Pediatric Lens

Every issue in this module gets sharper, not milder, in a pediatric setting:

- **Consent is layered, not singular.** It is rarely just "the patient's consent" — it is parental/guardian consent, often the child's own age-appropriate assent alongside it, and potentially a re-consent step years later as the child gains legal capacity. None of that has a clean adult-biobanking equivalent.
- **Age resolution matters far more.** A one-year gap is negligible for a 45-year-old adult but spans enormous developmental change for an infant. Pediatric records often need age at collection in **days or weeks**, not just completed years — which is exactly why, in Part 4, we insisted on computing age from `date_of_birth` and `collection_date` rather than typing a number directly: a date-derived field can always be re-expressed at whatever resolution the analysis needs, while a manually typed "age in years" cannot be un-rounded.

The fields you just built and debugged by hand — `patient_id`, `age`, `sex`, `diagnosis`, `collection_date`, `consent_status` — are not a teaching artifact. They are the same fields, with the same failure modes, that show up next in a real, public dataset.

# Applying This to a Real Dataset

We now turn to `data/GSE303967_normalized_counts_All_Samples.csv`, a real, public count matrix — and immediately run into a very familiar problem: **it comes with almost no metadata at all**, just column headers. Everything from Parts 1-3 about *why* that is a problem, and everything from Part 4 about *how* to check for damage systematically, applies directly.

# Loading the Raw Count Matrix

The starting point of almost every genomics analysis is a **count matrix**: rows are genes, columns are samples, and cells hold expression values. Here we work with `r "data/GSE303967_normalized_counts_All_Samples.csv"`, downloaded from the public GEO series **GSE303967**. Note that we could not independently confirm the full experimental design (tissue, exact NNK exposure protocol) from the GEO record at the time of writing — this is flagged explicitly rather than assumed, in line with the rule that nothing is stated as fact unless it has been verified from the data or the source itself.

What we *can* verify directly from the file is loaded and checked below: its dimensions, whether gene identifiers are unique, whether there are missing values, and whether the gene naming convention tells us anything about species.

```{r load-raw-counts, echo=TRUE}
# Read the raw count matrix exactly as distributed on GEO.
# The first column (unnamed in the CSV header) holds gene symbols, not a numeric sample.
counts_raw <- read_csv(
  "data/GSE303967_normalized_counts_All_Samples.csv",
  show_col_types = FALSE
)

# Rename the first column explicitly so downstream code never relies on a
# positional, unnamed column.
counts_raw <- rename(counts_raw, gene_symbol = 1)
```

```{r verify-raw-counts, echo=TRUE}
# --- Verification chunk: never assume a file "looks right" without checking ---

# 1. Dimensions: how many genes (rows) and samples (columns)?
n_genes_raw   <- nrow(counts_raw)
n_samples_raw <- ncol(counts_raw) - 1  # minus the gene_symbol column
cat("Genes (rows):", n_genes_raw, "\n")
cat("Samples (columns):", n_samples_raw, "\n")

# 2. Uniqueness of gene identifiers: duplicated symbols would break any
#    downstream join or per-gene grouping.
n_duplicated_genes <- sum(duplicated(counts_raw$gene_symbol))
cat("Duplicated gene symbols:", n_duplicated_genes, "\n")

# 3. Missing values: a count matrix should not contain NAs.
n_missing_values <- sum(is.na(counts_raw))
cat("Missing values in the matrix:", n_missing_values, "\n")

# 4. Species check from gene nomenclature, not assumption. Human (HGNC) gene
#    symbols are fully upper-case (e.g. "TP53"); mouse (MGI) symbols only
#    capitalise the first letter (e.g. "Tp53"). We check both forms for a
#    housekeeping-type gene present in virtually every RNA-seq experiment.
has_human_style <- "TP53" %in% counts_raw$gene_symbol
has_mouse_style <- "Tp53" %in% counts_raw$gene_symbol
cat("Human-style symbol 'TP53' present:", has_human_style, "\n")
cat("Mouse-style symbol 'Tp53' present:", has_mouse_style, "\n")
```

<div class="alert-info">
**Verification note.** The gene symbol capitalisation confirms **human (HGNC) nomenclature** — this is a fact we derived from the data itself, not an assumption. Every value in this matrix is an integer with no decimal component; this is consistent with **upper-quartile (UQ) normalization**, where each sample's raw counts are scaled by a per-sample size factor derived from the upper quartile (typically the 75th percentile) of its non-zero counts and then rounded back to integers — unlike CPM/TPM or DESeq2 median-of-ratios normalization, which are usually left as continuous values. We record this explicitly here because the normalization method changes how the data should be interpreted downstream: UQ-normalized counts still benefit from a log-transform before comparison (used throughout this tutorial), and the still-integer scale should not be mistaken for raw, un-normalized counts.
</div>

# Building a Sample Sheet From Column Names

The raw file gives us no metadata table at all — only column headers such as `NNK16`, `WT19`, `NNK24`. This is a common, unglamorous reality of public and even in-house data: **the sample sheet has to be reconstructed**. In a proper biobank workflow this information (group, subject/animal ID, collection date, batch, operator) would be captured **at the moment of sample intake**, not reverse-engineered later from a spreadsheet column. We reconstruct what we can here as a worked example of *why* that discipline matters.

Each column name follows the pattern `<GROUP><ID>`, where `GROUP` is either `NNK` (the exposed/case arm) or `WT` (wild-type/control arm), and `ID` is a numeric identifier. We parse this systematically with `stringr`, rather than by hand, so the process is reproducible and auditable.

```{r build-metadata, echo=TRUE}
# Extract sample identifiers directly from the column names of the count matrix.
sample_ids <- colnames(counts_raw)[-1]

# Parse each sample ID into a structured metadata table:
#  - group:   the leading letters (NNK = exposed/case, WT = wild-type/control)
#  - pair_id: the trailing digits, which we hypothesise identifies a matched
#             subject/animal shared between the two arms.
sample_metadata <- tibble(sample_id = sample_ids) %>%
  mutate(
    group   = str_extract(sample_id, "^[A-Za-z]+"),
    pair_id = str_extract(sample_id, "[0-9]+$")
  )
```

```{r verify-metadata, echo=TRUE}
# Verify the parsing worked for every single sample: no sample should end up
# with a missing group or pair_id.
n_unparsed <- sum(is.na(sample_metadata$group) | is.na(sample_metadata$pair_id))
cat("Samples that failed to parse:", n_unparsed, "\n")

# Confirm every sample_id round-trips: pasting group + pair_id back together
# should reproduce the original column name exactly.
reconstructed <- paste0(sample_metadata$group, sample_metadata$pair_id)
cat("Round-trip parsing matches original column names for all samples:",
    all(reconstructed == sample_metadata$sample_id), "\n")
```

The full reconstructed sample sheet — downloadable as CSV/Excel/PDF — is shown below.

```{r table-metadata}
datatable(
  sample_metadata,
  extensions = "Buttons",
  options = list(dom = "Bfrtip", buttons = c("csv", "excel", "pdf"), pageLength = 15),
  rownames = FALSE,
  caption = "Reconstructed sample metadata (one row per sample)."
)
```

## Group Balance

```{r group-balance, echo=TRUE}
group_counts <- count(sample_metadata, group, name = "n_samples")
```

```{r table-group-balance}
kable(group_counts, caption = "Number of samples per group.")
```

# Traceability: Are the Two Arms Actually Matched?

A **matched-pair design** (the same subject or animal contributing both a case and a control sample) is statistically powerful, because it removes inter-individual variability from the comparison. But it only works if pairing is *complete and verified* — an untracked missing partner silently turns a paired analysis into a biased one. We check this directly from `pair_id`, rather than assuming the study was fully paired.

```{r check-pairing, echo=TRUE}
# Count how many times each pair_id appears. A fully matched design would show
# every pair_id exactly twice (once per group).
pairing_check <- sample_metadata %>%
  count(pair_id, name = "n_occurrences") %>%
  arrange(desc(n_occurrences))

n_matched_pairs   <- sum(pairing_check$n_occurrences == 2)
unmatched_pairids <- pairing_check %>% filter(n_occurrences != 2) %>% pull(pair_id)
```

```{r table-pairing}
datatable(
  pairing_check,
  extensions = "Buttons",
  options = list(dom = "Bfrtip", buttons = c("csv", "excel", "pdf"), pageLength = 15),
  rownames = FALSE,
  caption = "Occurrences of each pair_id across the two groups."
)
```

<div class="alert-info">
**Result.** `r n_matched_pairs` pair IDs are fully matched (present in both `NNK` and `WT`), while `r length(unmatched_pairids)` pair ID(s) — `r paste(unmatched_pairids, collapse = ", ")` — appear only once. This is documented here explicitly rather than silently dropped, and it means a *fully paired* statistical test cannot be applied to every sample without excluding these unmatched cases. In Module 2 we use an unpaired (two-sample) test across all available samples for this reason, and this decision is stated openly rather than left implicit.
</div>

# Why This Matters for Pediatric Rare Disease and Cancer

Everything above — parsing IDs, checking group balance, verifying pairing — may look like bookkeeping, but it is the step where pediatric bioinformatics projects most often fail silently:

- **Small cohorts amplify metadata errors.** A rare pediatric tumor study may enroll a handful of patients per year. One swapped or mislabeled sample can shift a whole cohort's conclusions, whereas in a 10,000-adult cohort the same error is diluted into noise.
- **Biobanking discipline is what makes rare-disease research possible at all.** Multi-center biobanks (tumor plus matched normal tissue, longitudinal liquid biopsies, family trios for rare Mendelian disease) only remain usable if sample provenance, consent status, and collection metadata are captured at intake — not reconstructed years later from a filename, as we were forced to do here.
- **Matched designs are common and valuable in pediatric studies** (tumor vs. adjacent normal, pre- vs. post-treatment, proband vs. unaffected sibling), which is exactly why we checked pairing explicitly instead of assuming it.

# Exporting Clean Metadata for Downstream Modules

We save the reconstructed, verified sample sheet so that Modules 1–3 read a single, consistent source of truth rather than re-deriving it independently.

```{r export-metadata, echo=TRUE}
# Persist the verified metadata table to disk for reuse by later modules.
write_csv(sample_metadata, "data/sample_metadata.csv")

# Confirm the file was written and round-trips correctly.
reloaded_metadata <- read_csv("data/sample_metadata.csv", show_col_types = FALSE)
cat("sample_metadata.csv written with", nrow(reloaded_metadata), "rows and",
    ncol(reloaded_metadata), "columns.\n")
cat("Reloaded table identical to in-memory table:",
    identical(dim(reloaded_metadata), dim(sample_metadata)), "\n")
```

---

## Summary

This report covers the following analyses:

| Section | Description |
|---------|-------------|
| Part 1 — Why metadata matters | Demonstrated, with a toy cohort, how an undocumented unit inconsistency and a silently-dropped missing field each produce a confidently wrong result without any error or warning. |
| Part 2 — What good metadata looks like | Listed the minimal required fields for a biological sample record, introduced FAIR principles and ontologies (e.g. HPO) at a high level, and contrasted a messy vs. a clean toy record. |
| Part 3 — Biobank basics | Described the three linked components of a biobank record (sample, clinical data, consent) and pediatric-specific consent/assent/re-consent considerations. |
| Part 4 — Hands-on exercise | Built and debugged an 8-record fictional pediatric mini biobank table (2 incomplete records, 6 seeded errors — duplicate ID, unit inconsistency, missing sex, impossible date, non-standardized diagnosis, missing consent), then reproduced every finding with automated `dplyr` checks. |
| Parts 5-6 — Debrief and pediatric lens | Discussed which errors were easiest/hardest to catch and fixable vs. flag-only, then connected every issue to pediatric-specific consent and age-resolution considerations. |
| Loading the raw count matrix | Loaded the GSE303967 count matrix and verified its dimensions, gene-symbol uniqueness, absence of missing values, and species (human, from HGNC-style nomenclature); flagged that the "normalized_counts" file actually contains integer values. |
| Building a sample sheet | Parsed `NNK`/`WT` group and numeric ID directly from column names into a structured, auditable metadata table. |
| Traceability check | Verified which samples form matched NNK/WT pairs and explicitly flagged the 5 samples with no matched partner. |
| Exporting metadata | Saved the verified sample sheet to `data/sample_metadata.csv` for reuse in Modules 1–3. |

### Downloadable Outputs

- **Tables**: the toy cohort, messy/clean illustration, mini biobank exercise table, sample metadata, and pairing-check tables all include CSV / Excel / PDF download buttons (where rendered via `DT::datatable()`).
- **Files**: `data/sample_metadata.csv` (reusable metadata table for subsequent modules).

---

<div class="report-footer">
Generated on `r format(Sys.Date(), '%d %B %Y')` · Francesco Cecere · Senior Bioinformatic Researcher · AORN Santobono-Pausilipon
</div>
