---
title: "Module 3 — From Biomarkers to Clinical Outcome: Survival Analysis"
subtitle: "BBCC2026 Tutorial: Does This Candidate Gene Actually Matter for the Patient?"
author: |
  **Francesco Cecere**
  *Senior Bioinformatic Researcher*
  AORN Santobono-Pausilipon
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

<style>
body { font-size: 15px; }

h1 {
  border-bottom: 3px solid #2980b9;
  padding-bottom: 8px;
  color: #1a252f;
  margin-top: 32px;
}
h2 {
  border-bottom: 1px solid #aed6f1;
  padding-bottom: 4px;
  color: #2471a3;
  margin-top: 22px;
}
h3 { color: #1f618d; }

blockquote {
  background-color: #eaf4fd;
  border-left: 5px solid #2980b9;
  padding: 14px 20px;
  border-radius: 0 6px 6px 0;
  font-style: italic;
  color: #2c3e50;
  margin: 18px 0;
}
blockquote p { margin: 0; }

.value-boxes {
  display: flex;
  flex-wrap: wrap;
  gap: 16px;
  margin: 24px 0 32px 0;
}
.vbox {
  background-color: #2980b9;
  background: linear-gradient(135deg, #2980b9 0%, #1a5276 100%);
  color: #ffffff !important;
  border-radius: 10px;
  padding: 20px 28px;
  min-width: 150px;
  text-align: center;
  box-shadow: 0 3px 10px rgba(0,0,0,0.15);
  flex: 1;
  transition: transform 0.2s;
}
.vbox:hover { transform: translateY(-3px); }
.vbox.green  { background-color: #27ae60; background: linear-gradient(135deg, #27ae60 0%, #1e8449 100%); }
.vbox.red    { background-color: #e74c3c; background: linear-gradient(135deg, #e74c3c 0%, #a93226 100%); }
.vbox.purple { background-color: #8e44ad; background: linear-gradient(135deg, #8e44ad 0%, #6c3483 100%); }
.vbox .vbox-number, .vbox-number {
  font-size: 2.6em; font-weight: 700; line-height: 1.1; color: #ffffff !important;
}
.vbox .vbox-label, .vbox-label {
  font-size: 0.82em; margin-top: 5px; opacity: 0.92; letter-spacing: 0.02em; color: #ffffff !important;
}

.alert-sig {
  background: #d5f5e3;
  border-left: 5px solid #27ae60;
  padding: 12px 18px;
  border-radius: 0 6px 6px 0;
  margin: 12px 0;
}
.alert-notsig {
  background: #fdf2f8;
  border-left: 5px solid #a569bd;
  padding: 12px 18px;
  border-radius: 0 6px 6px 0;
  margin: 12px 0;
}
.alert-info {
  background: #fef9e7;
  border-left: 5px solid #f39c12;
  padding: 12px 18px;
  border-radius: 0 6px 6px 0;
  margin: 12px 0;
}

.dataTables_wrapper { margin: 10px 0 24px 0; font-size: 0.93em; }
table.dataTable thead th {
  background-color: #2980b9 !important;
  color: white !important;
}
.dt-buttons .btn { font-size: 0.82em; padding: 3px 10px; }

.report-footer {
  margin-top: 40px;
  padding-top: 16px;
  border-top: 1px solid #ccc;
  font-size: 0.85em;
  color: #777;
  text-align: center;
}

.tocify {
  background-color: #1a252f !important;
  border: none !important;
  border-radius: 8px;
  padding: 6px 0;
}
.tocify .tocify-item > a {
  color: #bdc3c7 !important;
  padding: 6px 14px;
  font-size: 0.88em;
  border-radius: 4px;
  transition: background 0.18s, color 0.18s;
  text-decoration: none;
}
.tocify .tocify-item > a:hover {
  background-color: #2c3e50 !important;
  color: #ffffff !important;
}
.tocify .tocify-item.active > a {
  background-color: #2980b9 !important;
  color: #ffffff !important;
  font-weight: 600;
  border-radius: 4px;
}
.tocify-subheader {
  background-color: #1a252f !important;
}
.tocify-subheader .tocify-item > a {
  padding-left: 26px !important;
  font-size: 0.82em !important;
  color: #7f8c8d !important;
}
.tocify-subheader .tocify-item.active > a {
  background-color: #21618c !important;
  color: #ecf0f1 !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  dpi       = 96
)
```

```{r libraries, include=FALSE}
library(tidyverse)   # dplyr, ggplot2, tidyr, readr, etc.
library(survival)    # Surv(), survfit(), survdiff() for Kaplan-Meier and the log-rank test
library(broom)       # tidy() converts a survfit object into a tidy tibble
library(plotly)      # interactive plots
library(DT)          # interactive downloadable tables
library(knitr)       # kable for simple inline tables
library(scales)      # axis formatting
```

# From Module 2

Module 2 produced a genome-wide, ranked list of candidate biomarkers (`results/differential_expression_results.csv`), flagged an unusually broad significant signal, and noted that several top genes look like markers of a specific cell type rather than confirmed carcinogenesis genes — a caveat carried forward, not resolved. This module asks the natural next question.

**Session plan (≈65 minutes):** why expression alone isn't enough (10 min) → Kaplan-Meier in plain language (15 min) → guided walkthrough on our real dataset (20 min) → light hands-on exploration (15 min) → closing pediatric lens (5 min).

# Part 1 — Why Expression Alone Isn't Enough

A gene can be textbook "significant" — a large fold-change, a tiny adjusted p-value, a clean volcano plot dot — and still be clinically useless. Statistical significance in an expression comparison answers one narrow question: *does this gene's average level differ between two groups in this dataset?* It says nothing about whether that difference has any bearing on what actually happens to a patient over time.

**Reframing the arc:** Module 2 found candidates by asking "is this gene behaving abnormally?" Module 3 asks a different, harder question about those same candidates: **"does it matter?"** — does knowing a patient's expression level for this gene tell us anything about their subsequent clinical course? Answering that requires linking expression back to an actual outcome measured over time, which is exactly what survival analysis is built for.

<div class="alert-info">
**A necessary disclosure before we go further.** As established explicitly in Module 2 (Part 2, the bridge to Module 0), **GSE303967 ships with no linked clinical follow-up data at all** — no survival time, no relapse status, nothing beyond a sample ID and an exposure-group label. A real biomarker-to-outcome analysis like this one is normally run on a dataset such as the NCI's **TARGET** pediatric cancer cohorts (via the GDC Data Portal), which do carry linked survival data. To keep this tutorial fully reproducible on Binder using the one real dataset we have throughout, the rest of this module uses the **real, actual expression values** from GSE303967 together with a **clearly labeled, simulated survival outcome** that we generate ourselves, purely to teach the Kaplan-Meier/log-rank methodology. Every simulated field is named with a `simulated_` prefix and is flagged again at the point it is created. **Nothing in this module should be read as a real finding about real patients** — the workflow is real; the outcome data is not.
</div>

# Part 2 — Kaplan-Meier in Plain Language

Forget formulas for a moment. A **Kaplan-Meier curve** answers one question over time: **"what percentage of patients are still alive (or still relapse-free) as of this many months since baseline?"** It starts at 100% (everyone is still event-free at time zero) and steps downward each time a patient in that group has the event — a death, a relapse, whatever the study defines as "the event." Patients who leave the study without the event happening yet (lost to follow-up, or the study simply ended) are **censored**: we know they were event-free up to that point, but not what happened after, and the curve accounts for that correctly rather than treating them as failures.

**How to read two curves plotted together:** if one group's curve stays consistently higher than the other's for most of the follow-up period, that group is doing better — a larger fraction of them are still event-free at any given time. **The bigger the vertical gap between the two curves, and the longer it persists, the stronger the apparent difference between groups.**

```{r toy-km-illustration, echo=TRUE, fig.height=5}
# A purely illustrative toy example — generic "Group A/B", not tied to any
# gene — just to practice reading a KM plot before using real expression data.
set.seed(1)
toy_data <- tibble(
  group = rep(c("Group A", "Group B"), each = 20),
  time  = c(rexp(20, rate = 0.03), rexp(20, rate = 0.08)),
) %>%
  mutate(time = pmin(time, 24), event = as.integer(time < 24))

toy_fit <- survfit(Surv(time, event) ~ group, data = toy_data)
toy_km <- tidy(toy_fit) %>% mutate(group = str_remove(strata, "group="))

p_toy <- toy_km %>%
  ggplot(aes(x = time, y = estimate, color = group)) +
  geom_step(linewidth = 1) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_color_manual(values = c("Group A" = "#2980b9", "Group B" = "#e74c3c")) +
  labs(x = "Time (months)", y = "% still event-free", color = NULL,
       title = "How to read a Kaplan-Meier plot (illustrative toy data)") +
  theme_minimal(base_size = 11)

ggplotly(p_toy)
```

The **log-rank test** is simply the formal statistical test for "are these two curves actually different, or could a gap this size have happened by chance alone?" — it produces a single p-value summarizing the whole curve comparison, not just one time point.

# Part 3 — Guided Walkthrough on Our Real Dataset

We now build exactly this kind of comparison using real GSE303967 expression values, split into "High" vs "Low" groups by a candidate gene from Module 2 — with the simulated outcome disclosed in Part 1 attached.

```{r load-data, echo=TRUE}
# Reload the real differential expression results, count matrix, and metadata.
de_results <- read_csv("results/differential_expression_results.csv", show_col_types = FALSE)
counts_raw <- read_csv("data/GSE303967_normalized_counts_All_Samples.csv", show_col_types = FALSE) %>%
  rename(gene_symbol = 1)
sample_metadata <- read_csv("data/sample_metadata.csv", show_col_types = FALSE)

# The index gene: the single top candidate from Module 2. Its REAL expression
# values are what we use throughout — only the outcome is simulated.
index_gene <- de_results %>% slice_head(n = 1) %>% pull(gene_symbol)
cat("Index gene (top Module 2 candidate):", index_gene, "\n")
```

```{r build-expression-table, echo=TRUE}
# Real, measured log2 expression of the index gene for every real sample.
index_expr <- counts_raw %>%
  filter(gene_symbol == index_gene) %>%
  pivot_longer(-gene_symbol, names_to = "sample_id", values_to = "count") %>%
  mutate(log2_count = log2(count + 1)) %>%
  select(sample_id, log2_count) %>%
  left_join(sample_metadata, by = "sample_id")

cat("Samples with expression data:", nrow(index_expr), "\n")
cat("Any missing expression values:", sum(is.na(index_expr$log2_count)), "\n")
```

```{r simulate-outcome, echo=TRUE}
# --- SIMULATED SURVIVAL OUTCOME — disclosed in Part 1, not a real finding ---
# We generate a synthetic time-to-event outcome per real sample from an
# exponential-hazard model where the hazard depends on this sample's real,
# z-scored index-gene expression: higher expression -> higher simulated
# hazard (shorter simulated_time_months, more likely simulated_event = 1).
set.seed(20260909)
z_index <- as.numeric(scale(index_expr$log2_count))
beta <- 0.7               # simulated effect size on the log-hazard scale
baseline_hazard <- 0.04   # simulated baseline hazard per month
simulated_hazard <- baseline_hazard * exp(beta * z_index)
true_event_time <- rexp(nrow(index_expr), rate = simulated_hazard)
follow_up_limit <- 36     # simulated administrative censoring at 36 months

survival_data <- index_expr %>%
  mutate(
    simulated_time_months = pmin(true_event_time, follow_up_limit),
    simulated_event       = as.integer(true_event_time <= follow_up_limit)
  )
```

```{r verify-simulated-outcome, echo=TRUE}
cat("Simulated events observed:", sum(survival_data$simulated_event),
    "out of", nrow(survival_data), "samples\n")
cat("Simulated censoring rate:",
    scales::percent(mean(survival_data$simulated_event == 0)), "\n")
cat("Median simulated follow-up (months):", round(median(survival_data$simulated_time_months), 1), "\n")
```

## Splitting Into High vs. Low Expression

```{r median-split, echo=TRUE}
# The simplest possible split: above vs. below the median real expression
# value for the index gene. Ties at the median are broken toward "Low" —
# the same convention used consistently by the Part 4 explorer widget below,
# so the two sections' median-split results agree exactly.
median_cut <- median(survival_data$log2_count)
survival_data <- survival_data %>%
  mutate(expr_group = if_else(log2_count > median_cut, "High", "Low"))

cat("Samples per group:\n")
print(count(survival_data, expr_group))
```

```{r fit-km, echo=TRUE}
# Fit the Kaplan-Meier estimator per group and tidy it with broom, exactly
# as t.test() results were tidied in Module 2 — same pattern, new model type.
km_fit <- survfit(Surv(simulated_time_months, simulated_event) ~ expr_group, data = survival_data)
km_tidy <- tidy(km_fit) %>% mutate(expr_group = str_remove(strata, "expr_group="))

logrank_test <- survdiff(Surv(simulated_time_months, simulated_event) ~ expr_group, data = survival_data)
logrank_p <- pchisq(logrank_test$chisq, df = 1, lower.tail = FALSE)
```

```{r verify-km, echo=TRUE}
cat("KM curve rows (event/censoring steps):", nrow(km_tidy), "\n")
cat("Log-rank p-value (", index_gene, ", median split):", signif(logrank_p, 4), "\n")
```

```{r plot-km, fig.height=6}
p_km <- km_tidy %>%
  ggplot(aes(x = time, y = estimate, color = expr_group)) +
  geom_step(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = expr_group), alpha = 0.12, color = NA) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_color_manual(values = c(High = "#e74c3c", Low = "#2980b9")) +
  scale_fill_manual(values = c(High = "#e74c3c", Low = "#2980b9")) +
  labs(x = "Simulated time (months)", y = "% simulated-event-free",
       color = paste(index_gene, "expression"), fill = paste(index_gene, "expression"),
       title = paste0(index_gene, " — High vs. Low expression (simulated outcome), log-rank p = ", signif(logrank_p, 3))) +
  theme_minimal(base_size = 11)

ggplotly(p_km)
```

<div class="alert-info">
**Keep the connection to Module 2 explicit.** This is the same `r nrow(survival_data)` real samples, split by the real expression of the same top candidate gene Module 2 flagged — the only new ingredient is the simulated outcome attached in Part 1. In a real study with real linked survival data (a TARGET-style cohort, for instance), this exact code — `survfit()`, `tidy()`, `survdiff()` — is what turns a Module-2-style candidate list into a statement about clinical relevance.
</div>

# Part 4 — Light Hands-On: Guided Exploration

Change the **gene** or the **split threshold** below and watch the curve update. The dropdown selects among the index gene (`r index_gene`, which the simulated outcome above was built from) and three genes drawn from the *opposite* end of Module 2's ranking (`p_adj` near 1, i.e. not flagged as candidates at all), each combined with three ways of splitting High vs. Low.

<div class="alert-info">
**A note on how this widget works, compared to Module 2's.** Module 2's explorer used `crosstalk` to *filter* rows of an already-drawn plot and table. A Kaplan-Meier curve cannot simply be filtered this way — changing the gene or the threshold changes *who is in which group*, which means the whole curve has to be **recomputed**, not just re-shown. So this widget instead pre-computes every gene × threshold combination and uses Plotly's built-in dropdown menu to switch which pre-drawn curve pair is visible. The experience is the same — pick an option, watch the result update instantly, no code to run — but the underlying mechanism is necessarily different.
</div>

```{r explorer-setup, echo=TRUE}
# Three "control" genes: the least significant genes from Module 2 (p_adj
# near 1), i.e. genes with no real evidence of differing between NNK and WT.
control_genes <- de_results %>% slice_tail(n = 3) %>% pull(gene_symbol)
explorer_genes <- c(index_gene, control_genes)
cat("Explorer genes:", paste(explorer_genes, collapse = ", "), "\n")
cat("Control gene p_adj values:", paste(signif(de_results %>% filter(gene_symbol %in% control_genes) %>% pull(p_adj), 3), collapse = ", "), "\n")

# Real expression for all four explorer genes, joined to the same simulated
# outcome defined once in Part 3 (the outcome does not change per gene).
explorer_expr <- counts_raw %>%
  filter(gene_symbol %in% explorer_genes) %>%
  pivot_longer(-gene_symbol, names_to = "sample_id", values_to = "count") %>%
  mutate(log2_count = log2(count + 1)) %>%
  select(gene_symbol, sample_id, log2_count) %>%
  pivot_wider(names_from = gene_symbol, values_from = log2_count) %>%
  left_join(survival_data %>% select(sample_id, simulated_time_months, simulated_event), by = "sample_id")
```

```{r explorer-compute, echo=TRUE}
# Three threshold options per gene: median split, and more extreme tertile /
# quartile splits that exclude an ambiguous "middle" group of samples.
thresholds <- tibble(
  threshold_label = c("Median split (50/50)",
                       "Tertile split (33/33, middle excluded)",
                       "Quartile split (25/25, middle excluded)"),
  lower_q = c(0.5, 1/3, 0.25),
  upper_q = c(0.5, 2/3, 0.75)
)

combos <- expand_grid(gene = explorer_genes, threshold_label = thresholds$threshold_label) %>%
  left_join(thresholds, by = "threshold_label") %>%
  mutate(combo_id = paste(gene, threshold_label, sep = " — "))

# For each gene x threshold combination: split into High/Low, fit the KM
# curve, and run the log-rank test. No custom function is defined beyond a
# small named helper reused via purrr::pmap, consistent with the rest of
# this tutorial's style.
fit_one_combo <- function(gene, lower_q, upper_q) {
  x <- explorer_expr[[gene]]
  low_cut <- quantile(x, lower_q)
  high_cut <- quantile(x, upper_q)
  split_data <- explorer_expr %>%
    mutate(expr_group = case_when(
      x <= low_cut  ~ "Low",
      x >= high_cut ~ "High",
      TRUE ~ NA_character_
    )) %>%
    filter(!is.na(expr_group))
  fit <- survfit(Surv(simulated_time_months, simulated_event) ~ expr_group, data = split_data)
  lr  <- survdiff(Surv(simulated_time_months, simulated_event) ~ expr_group, data = split_data)
  list(
    km   = tidy(fit) %>% mutate(expr_group = str_remove(strata, "expr_group=")),
    pval = pchisq(lr$chisq, df = 1, lower.tail = FALSE),
    n    = nrow(split_data)
  )
}

combo_results <- pmap(list(combos$gene, combos$lower_q, combos$upper_q), fit_one_combo)
all_km    <- map2_dfr(combo_results, combos$combo_id, ~ mutate(.x$km, combo_id = .y))
combo_pvals <- combos %>%
  mutate(logrank_p = map_dbl(combo_results, "pval"), n_samples = map_dbl(combo_results, "n")) %>%
  select(combo_id, gene, threshold_label, n_samples, logrank_p)
```

```{r verify-explorer, echo=TRUE}
cat("Gene x threshold combinations computed:", nrow(combo_pvals), "\n")
```

```{r table-explorer-pvalues}
kable(combo_pvals %>% mutate(logrank_p = signif(logrank_p, 3)),
      caption = "Log-rank p-value for every gene x threshold combination available in the widget below.")
```

```{r km-explorer-widget, fig.height=6}
# Build one Plotly figure containing every combination's two curves (High,
# Low), with only the first combination visible at load; the dropdown swaps
# which pair of traces is shown and updates the title with that combo's
# log-rank p-value.
combo_ids <- unique(all_km$combo_id)

fig <- plot_ly()
trace_lookup <- list()
for (cid in combo_ids) {
  for (grp in c("High", "Low")) {
    d <- all_km %>% filter(combo_id == cid, expr_group == grp)
    fig <- fig %>% add_trace(
      x = d$time, y = d$estimate, type = "scatter", mode = "lines",
      line = list(shape = "hv", color = if (grp == "High") "#e74c3c" else "#2980b9"),
      name = grp, legendgroup = cid,
      visible = (cid == combo_ids[1]), showlegend = (cid == combo_ids[1])
    )
    trace_lookup[[length(trace_lookup) + 1]] <- list(combo_id = cid, group = grp)
  }
}

dropdown_buttons <- map(combo_ids, function(cid) {
  visible_vec <- map_lgl(trace_lookup, ~ .x$combo_id == cid)
  p <- combo_pvals %>% filter(combo_id == cid) %>% pull(logrank_p)
  list(method = "update",
       args = list(list(visible = visible_vec, showlegend = visible_vec),
                    list(title = paste0(cid, " — log-rank p = ", signif(p, 3)))),
       label = cid)
})

fig %>% layout(
  updatemenus = list(list(y = 1.25, x = 0, xanchor = "left", buttons = dropdown_buttons)),
  title = paste0(combo_ids[1], " — log-rank p = ", signif(combo_pvals$logrank_p[1], 3)),
  yaxis = list(title = "% simulated-event-free", tickformat = ".0%", range = c(0, 1)),
  xaxis = list(title = "Simulated time (months)")
)
```

<div class="alert-info">
**What to notice.** Selecting `r index_gene` (any threshold) should consistently show a visible gap and a low log-rank p-value, since it is the gene the simulated outcome was actually built from. The three control genes should mostly show overlapping curves and non-significant p-values, as expected — **except** watch what happens if you check the p-value table above closely: at least one control gene reaches `p < 0.05` at one particular threshold, purely by chance. This is not a real finding — it is exactly the **threshold-shopping / multiple-comparisons risk** Module 2 already warned about: if you try enough gene-threshold combinations on data with no real relationship, one of them will look "significant" sooner or later. The fix is the same one used throughout this tutorial: decide your threshold *before* looking at the results, or correct for the number of comparisons made, rather than reporting whichever combination happened to look best.
</div>

```{r export-survival-results, echo=TRUE}
# Persist both the combination-level results and the simulated dataset used
# to produce them, clearly labeled, for anyone who wants to reuse them.
dir.create("results", showWarnings = FALSE)
write_csv(combo_pvals, "results/survival_logrank_results.csv")
write_csv(survival_data, "results/simulated_survival_dataset.csv")

cat("results/survival_logrank_results.csv written with", nrow(combo_pvals), "rows.\n")
cat("results/simulated_survival_dataset.csv written with", nrow(survival_data), "rows.\n")
```

# Part 5 — Closing Pediatric Lens

This module's Kaplan-Meier curves were fit on **`r nrow(survival_data)` samples**, with only **`r sum(survival_data$simulated_event)` simulated events** to work with. That is a genuinely small number of events for a survival analysis, and it is not an artifact of using simulated data — it is a realistic reflection of pediatric cohort sizes. A handful of events is exactly what makes Kaplan-Meier curves **statistically fragile**:

- **Wide confidence intervals, especially late in follow-up.** Notice in Part 3's plot how the shaded confidence band widens toward the right-hand side of the curve — every step down after an event happening removes a patient from the risk set, so the remaining estimate is based on fewer and fewer people, and its uncertainty grows accordingly.
- **Sensitivity to a handful of events.** With `r sum(survival_data$simulated_event)` events split across two groups, one or two events happening slightly earlier or later than they did could visibly shift where a curve sits — a single patient can measurably change the picture in a way that would be invisible in a hundreds-of-events adult cancer trial.
- **This is precisely why dedicated pediatric consortia exist.** Programs like the NCI's TARGET pool patients across many institutions specifically because no single pediatric oncology center sees enough cases of a given rare tumor type in a reasonable timeframe to produce a statistically stable survival curve on its own — the same small-numbers problem behind every module in this tutorial, now showing up in its clinical-outcome form.

Real biomarker validation in pediatric oncology has to hold both of these facts at once: a wide confidence interval is not a reason to ignore a promising signal, but it is a reason not to treat one small cohort's curve as a settled answer.

---

## Summary

This report covers the following analyses:

| Section | Description |
|---------|-------------|
| Part 1 — Why expression alone isn't enough | Reframed the Module 2 → Module 3 arc, and disclosed explicitly that GSE303967 has no real linked survival data, so a simulated outcome is used to teach the methodology. |
| Part 2 — Kaplan-Meier in plain language | Explained how to read a Kaplan-Meier curve and the log-rank test using a purely illustrative toy example, with no formulas. |
| Part 3 — Guided walkthrough | Built a simulated survival outcome tied to the real expression of Module 2's top candidate gene, fit and plotted Kaplan-Meier curves for a High/Low median split using `survival`/`broom`, and ran the log-rank test. |
| Part 4 — Guided exploration | Built a Plotly dropdown-driven explorer across 4 genes x 3 thresholds (12 combinations), and surfaced a genuine false-positive result among the control genes as a live illustration of threshold-shopping risk. |
| Part 5 — Closing pediatric lens | Used this analysis's own small event count to explain why Kaplan-Meier curves are statistically fragile in pediatric cohorts, and why dedicated multi-institution consortia like TARGET exist. |

### Downloadable Outputs

- **Tables**: the gene x threshold p-value table includes a `kable()` static table; the underlying data is also exported to CSV below.
- **Plots**: the toy KM illustration, the Part 3 KM plot, and the Part 4 dropdown explorer all include a PNG download button via the plotly toolbar (for the explorer, the currently visible combination).
- **Files**: `results/survival_logrank_results.csv` (log-rank p-value for every gene x threshold combination) and `results/simulated_survival_dataset.csv` (the simulated outcome data used throughout, clearly labeled).

---

<div class="report-footer">
Generated on `r format(Sys.Date(), '%d %B %Y')` · Francesco Cecere · Senior Bioinformatic Researcher · AORN Santobono-Pausilipon
</div>
